\begin{abstract}
  Hofmann (1999) introduced the functional programming language LFPL to characterize the functions computable in polynomial time using an affine type system.
  %
  LFPL enables a natural programming style, including nested recursion, and has inspired the development of type systems for automatic cost analysis, linear dependent type theories, and efficient memory management in functional programming languagesƒ.
  %
  Despite its prominence, there does not exist a self-contained presentation, let alone a full mechanization, of LFPL and its core metatheory.
  
  This article presents a modern account and mechanization of LFPL and its metatheory with the goal of being self-contained and accessible while presenting the strongest-known soundness and completeness results.
  %
  % TODO: Make macros for names (LFPL, LFPL+, AARA, etc).
  The soundness proof works with the language $\text{LFPL}^+$, which extends LFPL with additional language features, thereby providing a stronger theorem than pre-existing results.
  %
  The proof is novel, using a technique by Aehlig and Schwichtenberg (2002) to construct explicit polynomials that bound the cost of an $\text{LFPL}^+$ expression with respect to a big-step cost semantics.
  %
  The completeness proof shows that LFPL programs can simulate polynomial-time Turing machines while only relying on restricted forms of linear functions and lists.
  %
  It has the same structure as the original proof by Hofmann (2002) but greatly simplifies the core argument with a novel stack-like data structure, implemented with first-class functions and lists.
  %
  The mechanization includes the full soundness and completeness proofs and serves as one of the first case studies of mechanized metatheory in the recently-developed proof assistant Istari.
  %
  Takeaway?
\end{abstract}
