\section{Polynomial-Time Completeness}

Now that we have a denotational semantics, we can concretely state what it means for \lfpl to be complete with respect to polynomial-time computation. We focus on polynomial-time computations on the natural number. Since $\tpsem{\tplist{\tpunit}} = \{\ast\}^*$ is the set of finite-length strings from a unary alphabet, it is can be viewed as the set of unary-encoded natural numbers. However, we could just as easily have used the binary representation $\tpsem{\tplist{\tpsum{\tpunit}{\tpunit}}}$ for $\mathbb N$. So, instead of choosing a particular encoding, we can let the ``digits'' of the number be encoded by any type $A$ such that $\dfree{A}$. We require $\dfree{A}$ because plenty of polynomial-time computable functions (even non-size-increasing ones) duplicate the digits of their input, which is impossible in \lfpl if $A$ contains diamonds.

\begin{theorem}[Size-Restricted Polynomial-Time Completeness]
\label{thm:completeness-main}
Let $A$ be a type such that $\dfree{A}$. For every polynomial-time computable function $f : \tpsem{A}^* \to \tpsem{A}^*$ such that $|f(x)| \le |x|$ for all $x \in \tpsem{A}^*$, there exists a closed term $M : \tparr{\tplist{A}}{\tplist{A}}$ such that $\tmsem{M}{\envnil} = f$.
\end{theorem}

This theorem is originally proven in \cite{hofmannStrengthNonsizeIncreasing2002}. The constraint that $f$ does not increase the length of its input is necessary because of the non-size-increasing property; a closed \lfpl function on unit lists has no source of diamonds other than the elements of the input list, so the output list cannot have more diamonds (and therefore more elements) than the input list. As we will see later in \pref{sec:completeness-general-thm}, this constraint is relatively superficial and becomes unnecessary if we don't enforce that the output type of $M$ is an \lfpl list.

Proving \pref{thm:completeness-main} boils down to simulating the polynomial-time Turing machine which computes $f$. This immediately presents two challenges. Say that $P : \mathbb N \to \mathbb N$ is the polynomial function encoding how many steps the Turing machine takes to compute $f$. The first challenge is to simulate the tape of the machine, which for an input list of size $n$ could contain up to $P(n)$ additional values by the time the machine halts. We will only have $n$ diamonds from the input list, yet somehow we have to store up to $P(n)$ additional values; clearly a list will not work for this, since every element of a list must be accompanied by a diamond.

The second challenge is to simulate running the Turing machine for $P(n)$ steps. At first glance, it seems that \lfpl is only really suited for performing $kn$ iterations given $n$ diamonds and some constant $k$. This can be acheived for $k = 1$ by iterating down a list and building it back up during the iteration process, and then it is easy to repeat this for any constant $k$ number of times. This is useful but offers no real insight into how to perform, say, $n^2$ iterations.



\subsection{The Bounded Stack Data Structure}
\label{sec:completeness-bounded-stack-definition}

We first address the challenge of simulating the tape of the Turing machine. A common approach to this is to use a data structure consisting of a head element and two stacks, each representing one side of the tape relative to the head element \cite{Okasaki99}. Concretely, in \lfpl, if we use the type $A$ for tape symbols, then the type of our tape data structure would be $\tptens{\tplist{A}}{\tptens{A}{\tplist{A}}}$. The first element represents the section of the tape to the left of the head, the second element represents the symbol at the head of the tape, and the third element represents the right section of the tape. Treating elements of $L(A)$ like stacks, we can shift the head to the left by popping a symbol from the left list and pushing the current head symbol to the right list.

Of course, we cannot actually use the above data structure; given $n$ diamonds, it can only hold $n + 1$ values. Instead, we construct a type $S(A)$ (which may depend on $P$) that supports a stack-like interface, at least as long as it doesn't have to hold more than $P(n)$ elements. Then, we can use something of type $\tptens{S(A)}{\tptens{A}{S(A)}}$ to encode the tape.

We have actually already seen the first key insight into how to construct such a type in \pref{example:susp}. This shows us that we really don't need any diamonds to store data, we just need them to interact with it. Using this idea, we can design our data structure to remain fully suspended by default, requiring zero diamonds to store it. The push and pop functions will have to temporarily borrow some diamonds to interface with the suspended data.

The second key insight is that we do not need to un-suspend the entire data structure to interface with it. This is good, because if the stack held $P(n)$ values, we would still need $P(n)$ diamonds to un-suspend the whole thing during the push and pop operations. Since both push and pop work at the front of the stack, we just need to be able to un-suspend enough of our data structure to expose the front, which will save on the amount of diamonds we need for these operations. This amount may of course depend on $P$ and $n$. With these two insights in mind, we lay out our desired interface.

\begin{definition}[Bounded Stack Interface]
Given $k \in \mathbb N$ and an element type $A$, a $k$-stack implementation is a type $S$ together with the following closed terms:
\begin{itemize}
  \item $\code{empty} : S$
  \item $\code{push} : \tparr{(\tplist{\tpunit})^k}{\tparr{\tptens{A}{S}}{\tptens{(\tplist{\tpunit})^k}{\tptens{S}{(\tpsum{A}{\tpunit})}}}}$
  \item $\code{pop} : \tparr{(\tplist{\tpunit})^k}{\tparr{S}{\tptens{(\tplist{\tpunit})^k}{\tptens{S}{(\tpsum{\tpunit}{A})}}}}$
\end{itemize}
\end{definition}

By $(\tplist{\tpunit})^k$ we mean the type of $k$-tuples $\tplist{\tpunit} \otimes \cdots \otimes \tplist{\tpunit}$. We use it here as a source of diamonds; the lists contain no information, but the diamonds stored alongside each element will be used to un-suspend parts of the data structure. The reason for using a $k$-tuple of lists instead of just one big list becomes apparent in the proof of \pref{lem:stack-inductive}, where the use of such a tuple greatly simplifies the implementation.

Intuitively, both $\code{push}$ and $\code{pop}$ take in some diamonds and the return all of them as their first return value. They both also return a stack and a sum as their second and third return values. The left branch of the sum is intended to signal failure of the operation (in which case the input stack is returned unmodified), and the right branch signals success (in which case the returned stack has been successfully modified). For $\code{push}$, which takes in an element and a stack, failure means that the input stack is full, and so the input element is returned to us. For $\code{pop}$, which just takes in a stack, failure means the input stack is empty. If the input stack has at least one element, it is returned to us via the right injection.

This interface is inspired by the array-like data structure used in Hofmann's original completeness proof \cite{hofmannStrengthNonsizeIncreasing2002}. The key difference is that his array data structure has read and write operations requiring an index, given as a binary \lfpl number (i.e. $\tplist{\tpsum{\tpunit}{\tpunit}}$). The extra work of dealing with this index significantly complicates the implementation of the Turing machine's tape in his proof. Arrays have their advantages over stacks in many applications, where the user may want to access elements anywhere in the data structure, not just at the front. However, a stack-like interface is simpler, easier to implement, and is sufficient for representing a Turing machine's tape, so we believe it is the better choice of data structure for the completeness theorem.

To formalize our intuition of how $\code{push}$ and $\code{pop}$ should behave, we use the denotational semantics. We first define the element $m_{n,k} \in \tpsem{(\tplist{\tpunit})^k}$ to be the tuple $(\ell, \ldots, \ell)$, where $\ell$ is the unique element of $\tpsem{\tplist{\tpunit}}$ such that $|\ell| = n$. This way, an \lfpl term representing $m_{n,k}$ contains exactly $nk$ diamonds.

\begin{definition}[Bounded Stack Correctness]
\label{def:stack-correctness}
Let $k \in \mathbb N$. Suppose $(S, \code{empty}, \code{push}, \code{pop})$ is a $k$-stack implementation with element type $A$. Given a function $B : \mathbb N \to \mathbb N$, we say this implementation is correct and bounded by $B$ if the following holds for all $n \in \mathbb N$:
\begin{itemize}
  \item There exists a unary relation $V_n \subseteq \tpsem{S}$. This is intended to represent the valid states of the data structure.
  \item There exists a function $I_n : \tpsem{S} \to \tpsem{A}^*$. This is intended to represent the list of items contained in the stack.
  \item Let $s = \tmsem{\code{empty}}{\ctxnil}$. Then, $V_n(s)$ holds and $I_n(s) = []$.
  \item Let $f = \tmsem{\code{push}}{\ctxnil}$. Then, for all $x \in \tpsem{A}$ and $s \in \tpsem{S}$ such that $V_n(s)$ holds, we have:
  \begin{itemize}
    \item If $|I_n(s)| = B(n)$, then $f(m_{n,k})(x, s) = (m_{n,k}, s, (1, x))$.
    \item If $|I_n(s)| < B(n)$, then $f(m_{n,k})(x, s) = (m_{n,k}, s', (2, \ast))$, where $V_n(s')$ holds and $I_n(s') = x :: I_n(s)$.
  \end{itemize}
  \item Let $f = \tmsem{\code{pop}}{\ctxnil}$. Then, for all $s \in \tpsem{S}$ such that $V_n(s)$ holds:
  \begin{itemize}
    \item If $I_n(s) = []$, then $f(m_{n,k})(s) = (m_{n,k}, s, (1, \ast))$.
    \item If $I_n(s) = x :: \ell$, then $f(m_{n,k})(s) = (m_{n,k}, s', (2, x))$, where $V_n(s')$ holds and $I_n(s') = \ell$.
  \end{itemize}
\end{itemize}
\end{definition}

The intended use of the data structure is to choose $n \in \mathbb N$ and always pass $m_{n,k}$ to the $\code{push}$ and $\code{pop}$ operations. In other words, we commit to a value of $n$ at runtime and always provide $nk$ diamonds to the stack operations. \pref{def:stack-correctness} says that, in this use case, the stack can hold at most $B(n)$ items. Behavior is otherwise undefined. Note that whenever $\code{push}$ and $\code{pop}$ receive $m_{n,k}$ as input, they return $m_{n,k}$ as the first element of their output. This way, no diamonds are ever permanently used, only temporarily borrowed, and thus available for future stack operations.

The definition only makes claims about valid stacks, i.e. $s \in \tpsem{S}$ such that $V_n(s)$ holds. Naturally, we insist that $\code{empty}$ should always be valid, and that $\code{push}$ and $\code{pop}$ should preserve validity. This extra layer of complexity is necessary; many of the implementations we provide in \pref{sec:completeness-bounded-stack-implementation} will require stack types $S$ which contain elements that would cause the $\code{push}$ and $\code{pop}$ operations to fail their specification. The extension of \lfpl with dependent types studied in \cite{atkeyPolynomialTimeDependent2024} could be helpful here, potentially allowing us to internalize \pref{def:stack-correctness} into the types themselves.

The intuition for $\code{push}$ is that, given an element to push $x \in \tpsem{A}$ and a valid input stack $s \in \tpsem{S}$, we case on whether $s$ is full, i.e. $|I_n(s)| = B(n)$. If it's full, we just return $s$ and $x$. Otherwise, we return some new valid stack $s'$ such that the items of $s'$ are $x$ followed by the items of $s$. The intuition for $\code{pop}$ is quite similar, casing on whether $I_n(s)$ is $[]$ or $x :: \ell$ and proceeding accordingly.

To summarize, a correct $k$-stack bounded by $B : \mathbb N \to \mathbb N$ is a data structure which, given the user's choice of some number $n \in \mathbb N$ at runtime, behaves like a standard stack in a functional programming language but requires temporary access to $nk$ diamonds and has a storage limit of $B(n)$ elements.



\subsection{Bounded Stack Implementations}
\label{sec:completeness-bounded-stack-implementation}

We now provide a concrete implementation of a bounded stack and several ways to build better implementations on top of pre-existing ones. These can ultimately be combined to give us a stack bounded by our choice of a polynomial $P$, which is exactly what we want for simulating the tape of a polynomial-time Turing machine. While most of the proofs in this section focus on the high-level intuition and avoid writing or reasoning about concrete \lfpl programs, everything is formalized in the mechanization (see \pref{sec:mechanization}).

\begin{lemma}[Constant Stack Construction]
\label{lem:stack-constant}
For any element type $A$ and $c \in \mathbb N$, there exists a correct $0$-stack implementation which is bounded by $B(n) = c$.
\end{lemma}
\begin{proof}
We take the implementation type to be $S = (1 + A)^c$. Intuitively, a left injection represents an open space in the stack, and a right injection represents a stored element. A stack $(x_1, \ldots, x_c) \in \tpsem{S}$ is considered valid when there is some index $0 \le j \le c$ such that $x_i = (1, \ast)$ for $1 \le i \le j$, and $x_i$ is a right injection for $j < i \le c$. In other words, there should be an uninterrupted chunk of open slots at the front of the stack, and the rest of the slots afterwards should be filled.

To find the head of the stack, find the first $i$ such that $x_i = (2, a)$ for some $a \in \tpsem{A}$; the value $a$ is the head element. If $i$ is $1$, then the stack is full. If no such $i$ exists, the stack is empty. Since $c$ is a known constant, this procedure of locating the stack head does not require \lfpl iteration, so indeed no diamonds are required for the $\code{push}$ and $\code{pop}$ operations.
\end{proof}

The main difficulty is obtaining stacks with non-constant bounds. Surprisingly, the proof of \pref{lem:stack-constant} contains the right intuition for how to do that. The following lemma is essentially the non-constant version of \pref{lem:stack-constant}; instead of storing things in a fixed-size tuple, we store them in a list. That list can be ``suspended'' as seen in \pref{example:susp} so that it doesn't require any diamonds to store.

\begin{lemma}[Inductive Stack Construction]
\label{lem:stack-inductive}
Fix an element type $A$ and some $k \in \mathbb N$. Suppose $(S, \code{empty}, \code{push}, \code{pop})$ is a $k$-stack implementation with element type $A$ that is correct and bounded by $B : \mathbb N \to \mathbb N$. Then, there is a $(k + 1)$-stack implementation with element type $A$ that is correct and bounded by $B'(n) = n B(n)$. 
\end{lemma}
\begin{proof}
We choose $S' = \tparr{\tplist{\tpunit}}{\tplist{S}}$, the type of suspended lists with element type $S$ (we will these elements ``sub-stacks''), to be our implementation type. Let $\ell$ be the unique element of $\tpsem{\tplist{\tpunit}}$ with $|\ell| = n$. To unsuspend a stack $s \in \tpsem{S'}$ we just call $s(\ell)$. We say $s$ is valid whenever $s(\ell) = s_1 :: \cdots :: s_n :: []$, where $s_i$ is valid for all $1 \le i \le n$ and, much like in the proof \pref{lem:stack-constant}, there exists some index $0 \le j \le n$ such that $s_i$ is empty for $1 \le i < j$, $s_j$ is unconstrained, and $s_i$ is full for $j < i \le n$. In other words, there should be an uninterrupted chunk of empty sub-stacks at the front of the stack, followed by a nonempty sub-stack, and then all remaining sub-stacks should be full. The items $I'(S')$ are simply the item lists $I(s_i)$ of each sub-stack appended together.

It remains to correctly implement the three stack operations. Let's call them $\code{empty}'$, $\code{push}'$, and $\code{pop}'$. Implementing $\code{empty}'$ is quite straightforward; it is simply the suspended empty list, i.e. the function of type $\tparr{\tplist{\tpunit}}{\tplist{S}}$ which always returns the empty list.

When implementing $\code{push}'$, we are given as our source of diamonds an input of type $(\tplist{\tpunit})^{k+1} = \tptens{\tplist{\tpunit}}{(\tplist{\tpunit})^k}$, so we can use the first element of this tuple, call it $\ell$, to un-suspend the input stack $s \in \tpsem{S'}$ and obtain a list of sub-stacks $s(\ell) = s_1 :: \cdots :: s_n :: []$. Starting from the last element of this list, $s_n$, attempt pushing to it using the function $\code{push}$, which is correct by assumption. If this succeeds, we are done. Otherwise, $s_n$ must be full, so we should try $s_{n-1}$. We can do this all the way up to $s_1$; if push still fails on $s_1$ then the whole stack must be full. This process can be implemented using \lfpl's list recursor. Afterwards, we use \pref{example:susp} to re-suspend the modified list and return the results.

The implementation of $\code{pop}'$ is similar, but instead we start from $s_1$. If popping $s_1$ fails, it must be empty, so we should try $s_2$. We go all the way down the list, and if popping $s_n$ still fails, then the whole stack must be empty.
\end{proof}

Hofmann's original proof in \cite{hofmannStrengthNonsizeIncreasing2002} contains an analogue to \pref{lem:stack-inductive} for his array data structure, but the implementation is significantly more difficult due to the presence of indices. The construction is similar to ours; he chooses $S' = \tparr{\tplist{\tpunit}}{\tplist{S}}$, where $S$ is now assumed to be the implementation type of a correct array data structure, so $S'$ is the type of suspended lists of sub-arrays. Given an index between $0$ and $n B(n)$, the read and write operations need to figure out which sub-array corresponds to that index, and furthermore which index to use for the operations on that sub-array. This involves implementing the binary divmod operation in LFPL, greatly complicating the implementation and proof of its correctness.

Now that we've established the main result of this subsection, all it takes is a straightforward induction to combine the above two lemmas to produce stacks with monomial bounds.

\begin{lemma}[Monomial Stack Construction]
\label{lem:stack-monomial}
Given $c,k \in \mathbb N$, there exists a $k$-stack implementation with element type $A$ that is correct and bounded by $B(n) = cn^k$.
\end{lemma}

Since every polynomial $P : \mathbb N \to \mathbb N$ is eventually bounded by some monomial $B(n) = cn^k$, it is technically enough to stop here, which is what Hofmann's original proof does \cite{hofmannStrengthNonsizeIncreasing2002}. However, this introduces an annoying edge case when $n = 0$ because $P(0)$ could be nonzero but $B(0) = 0$ for any choice of $c$ and $k$. We suspect the complexity of the array data structure makes it not worth the trouble to avoid this problem. Due to the simplicity of our stack data structure, this annoyance can be sidestepped and we can combine monomially-bounded stacks to get polynomially-bounded stacks with a few more relatively straightforward implementations.

\begin{lemma}[Weakened Stack Construction]
\label{lem:stack-weakened}
Fix an element type $A$ and some $k \in \mathbb N$. Suppose $(S, \code{empty}, \code{push}, \code{pop})$ is a $k$-stack implementation with element type $A$ that is correct and bounded by $B : \mathbb N \to \mathbb N$. Then, there is a $(k + 1)$-stack implementation with element type $A$ that is correct and bounded by $B$. 
\end{lemma}
\begin{proof}
This construction is relatively obvious; if $\code{push}$ and $\code{pop}$ need only $kn$ diamonds, then instead having to accept $(k + 1)n$ diamonds is no problem. We can reuse the implementation type $S$ as well as the code for $\code{empty}$. For the new stack operations $\code{push}'$ and $\code{pop}'$, they simply make use of the fact that $m_{n,k+1} = (\ell,m_{n,k})$ for some $\ell \in \tpsem{\tplist{\tpunit}}$. They just ignore $\ell$ and call $\code{push}$ and $\code{pop}$ (respectively) with $m_{n,k}$.
\end{proof}

The following lemma, which shows how to combine two stacks into one stack bounded by the sum of their bounds, is the main tool by which we are able to construct stacks with arbitrary polynomial bounds from stacks with monomial bounds. It is straighforward to do this construction for our stack data structure, but Hofmann's array data structure presents difficulties here similar to the ones discussed after \pref{lem:stack-inductive}. We suspect this is why he stopped at monomially-bounded arrays in \cite{hofmannStrengthNonsizeIncreasing2002}.

\begin{lemma}[Additive Stack Construction]
\label{lem:stack-additive}
Fix an element type $A$ and some $k \in \mathbb N$. For $i \in \{1,2\}$, suppose $(S_i, \code{empty}_i, \code{push}_i, \code{pop}_i)$ is a $k$-stack implementation with element type $A$ that is correct and bounded by $B_i : \mathbb N \to \mathbb N$. Then, there is a $k$-stack implementation with element type $A$ that is correct and bounded by $B(n) = B_1(n) + B_2(n)$. 
\end{lemma}
\begin{proof}
We take the implementation type to be $S = \tptens{S_1}{S_2}$. A stack $(s_1, s_2) \in \tpsem{S}$ is considered valid when $s_1$ and $s_2$ are valid, and $s_1$ is only nonempty when $s_2$ is full. Intuitively, $s_1$ contains the first section of our stack and $s_2$ contains the second. The operations are very similar to that of \pref{lem:stack-constant} with $c = 2$. To pop the stack, we first try popping $s_1$. If that fails, it must've been empty, so we pop $s_2$. To push some $x \in \tpsem{A}$ onto this stack, we first try pushing to $s_2$. If that fails, it means $s_2$ is full, so we push to $s_1$.
\end{proof}

\begin{lemma}[Polynomial Stack Construction]
\label{lem:stack-polynomial}
Given a polynomial $P : \mathbb N \to \mathbb N$ with degree $k$, there exists a $k$-stack implementation with element type $A$ that is correct and bounded by $P$.
\end{lemma}
\begin{proof}
Write $P(n)$ as $\sum_{i=0}^k c_i n^i$. For each $0 \le i \le k$, we can use \pref{lem:stack-monomial} to obtain an $i$-stack with bound $B_i(n) = c_i n^i$. Then, we can repeatedly use \pref{lem:stack-weakened} to weaken all of them into $k$-stacks with the same bounds. Finally, we can repeatedly use \pref{lem:stack-additive} to obtain the desired $k$-stack whose bound is the sum of the bounds of the weakened monomial stacks.
\end{proof}



\subsection{Polynomial Iteration}
\label{sec:completeness-poly-iter}

We now address the challenge of how to simulate running the transition function of a $P$-time Turing machine for $P(n)$ steps when given an input list of length $n$. Additionally, we need to figure out how to simulate the transition function itself. Fortunately, both of these tasks require comparably simpler solutions to the issue of polynomial storage presented in the previous section. To iterate a function on itself $P(n)$ times, we use the next two lemmas, which are inspired by Proposition 4.1 in \cite{hofmannLinearTypesNonsizeincreasing1999}.

\begin{lemma}[Linear Iteration]
\label{lem:linear-iterator}
Let $f : \tparr{\tptens{A}{\tplist{\tpunit}}}{\tptens{A}{\tplist{\tpunit}}}$ be a closed term. Then, there exists a closed term $f^{\sharp} : \tparr{\tptens{A}{\tplist{\tpunit}}}{\tptens{A}{\tplist{\tpunit}}}$ such that $\tmsem{f^{\sharp}}{\envnil} (x, n) = (\tmsem{f}{\envnil})^{|n|}(x, n)$.
\end{lemma}
\begin{proof}
Consider the following pseudocode:
\begin{lstlisting}
g : L(1) -o A * L(1) -o A * L(1)
g = lam m . rec m
| nil => lam s . s
| cons (d, u, r) => lam s .
    letp (x, n) = s in f (r (x, cons (d, u, n)))
\end{lstlisting}
We can then define $f^{\sharp} = \tmlam{s}{\tmletp{s}{x}{n}{\tmapp{\tmapp{g}{n}}{\tmpair{x}{\tmnil}}}}$.
\end{proof}

\begin{lemma}[Polynomial Iteration]
\label{lem:polynomial-iterator}
Let $f : \tparr{\tptens{A}{\tplist{\tpunit}}}{\tptens{A}{\tplist{\tpunit}}}$ be a closed term and $P : \mathbb N \to \mathbb N$ be a polynomial. Then, there exists a closed term $f^{\sharp P} : \tparr{\tptens{A}{\tplist{\tpunit}}}{\tptens{A}{\tplist{\tpunit}}}$ such that:
$$\tmsem{f^{\sharp P}}{\envnil} (x, n) = (\tmsem{f}{\envnil})^{P(|n|)}(x, n)$$
\end{lemma}
\begin{proof}
To get monomials, we can use the $\sharp$ operation from \pref{lem:linear-iterator} multiple times, e.g. $\tmsem{(f^{\sharp})^{\sharp}}{\envnil} (x, n) = (\tmsem{f}{\envnil})^{|n|^2}(x, n)$. To add these monomials together to obtain any polynomial, we can just use function composition, since $g^m \circ g^k = g^{m + k}$ for any $g : X \to X$ and $m,k \in \mathbb N$.
\end{proof}

Finally, we address the task of encoding the transition function of Turing machine. Suppose are dealing with a tape alphabet that includes one blank symbol $a_0$ and the rest of the symbols are drawn from the set $\tpsem{A}$, where $A$ is a type such that $\dfree{A}$. Additionally, suppose set of internal states of the Turing machine includes one halting state $q_0$ and the rest of the states are drawn from the set $\tpsem{Q}$, where again $\dfree{Q}$. Then, the transition function $f$ of the Turing machine has the following signature:
$$\tpsem{Q} \times (\{ a_0 \} \sqcup \tpsem{A}) \to (\{ q_0 \} \sqcup \tpsem{Q}) \times (\{ a_0 \} \sqcup \tpsem{A}) \times \{ \Leftarrow, \Rightarrow \}$$

If $(q', a', d) = f(q, a)$, the intent is that $q$ is the current state of the Turing machine, $a$ is the symbol under the tape head, $q'$ is the new state after taking a step, $a'$ is the symbol to overwrite $a$ with, and $d$ is the direction (left or right) to shift the head. So, the type of the \lfpl term encoding $f$ should be:
$$\tparr{\tptens{Q}{(\tpsum{\tpunit{}}{A})}}{\tptens{(\tpsum{\tpunit{}}{Q})}{\tptens{(\tpsum{\tpunit{}}{A})}{(\tpsum{\tpunit}{\tpunit})}}}$$

But can we actually inhabit this type with a term that has the correct semantics? Fortunately, yes, given the following two lemmas that show \lfpl is complete with respect to computations only involving \dfree{\hspace{-0.75em}} types.

\begin{lemma}
Suppose $A$ is a type such that $\dfree{A}$. Then, for every $a \in \tpsem{A}$, there exists a closed term $M_a : A$ such that $\tmsem{M_a}{\envnil} = a$.
\end{lemma}

\begin{lemma}
Suppose $A$ and $B$ are types such that $\dfree{A}$ and, for every $b \in \tpsem{B}$, that there exists a closed term $M_b : B$ such that $\tmsem{M_b}{\envnil} = b$. Then, for every $f : \tpsem{A} \to \tpsem{B}$, there exists a closed term $M_f : \tparr{A}{B}$ such that $\tmsem{M_f}{\envnil} = f$.
\end{lemma}

Both of these are straightforward to prove by rule induction on $\dfree{A}$. By combining the two, we can obtain a closed term $M_f$ encoding the transition function $f$ of our Turing machine.



\subsection{Simulating the Turing Machine}
\label{sec:completeness-general-thm}

Now that we have dealt with the main challenges, it only remains to put everything together. For the rest of this section, assume we are dealing with a computable function $f : \tpsem{A}^* \to \tpsem{A}^*$, computed by a Turing machine on a tape with alphabet $\{a_\text{blank}\} \sqcup \tpsem{A}$, with internal state set $\{q_\text{final}\} \sqcup \tpsem{Q}$ and transition function $g$. The Turing machine always halts in $P(n)$ steps on an input of length $n$, where $P : \mathbb N \to \mathbb N$ is a polynomial of degree $k$.

Our goal is to construct a closed term $M : \tplist{A} \to \tplist{A}$ such that $\tmsem{M}{\envnil} = f$. Assuming an input of $\ell \in \tpsem{A}^*$ with length $n = |\ell|$, the body of the function for $M$ will only have access to $n$ diamonds. We are going to need a tape data structure that can hold up to $n + P(n)$ values while only requiring $n$ diamonds to use, since the input is written on the tape and (in the worst case) the Turing machine could just write a new value outside of the input section of the tape at every step.

Since the Turing machine receives its input as a single list, giving us a pool of diamonds in the form $\tplist{\tpunit}$, yet stack operations require diamonds in the form $(\tplist{\tpunit})^{k + 1}$, we need a way to losslessly convert between diamonds collected into a list of type $\tplist{\tpunit}$ and diamonds in a divided form $\tptens{(\tplist{\tpunit})^{k + 1}}{\tplist{\tpunit}}$. The reverse direction is straightforward; just append all the lists together into one. The forward direction amounts to implementing unary division by the constant $k + 1$ in \lfpl. This is nontrivial but ultimately follows a standard way one might implement $\code{divmod}_k : \mathbb N \to \mathbb N \times \mathbb N$ in a standard functional programming language, where $\code{divmod}_k(n)$ returns the quotient and remainder of $n$ by $k + 1$ (thus avoiding any division by zero issues).

Consider the polynomial $Q(m) = (k + 1)(m + 1) + P((k + 1)(m + 1))$ which has degree at most $k + 1$. Let $(m, r) = \code{divmod}_k(n)$ so that $0 \le r \le k$ and $n = m(k + 1) + r$. Then, $(k + 1)(m + 1) = m(k + 1) + (k + 1) > m(k + 1) + r = n$, implying $Q(m) \ge P(n) + n$. Therefore, we can apply \pref{lem:stack-polynomial} to $Q$ to obtain a stack implementation for element type $\tpsum{\tpunit{}}{A}$, say with underlying type $S$. These stacks hold up to $Q(m)$ values assuming we have $m(k + 1)$ diamonds available (which we do; $m(k + 1) < n$), and therefore at least $P(n)$ values. So, as discussed earlier, we can model the tape with a data structure of type $\tptens{S}{\tptens{(\tpsum{\tpunit{}}{A})}{S}}$, where shifting the tape head is implemented using the stack operations of $S$.

This is another place where our stack data structure simplifies the argument from Hofmann's array-based tape implementation in \cite{hofmannStrengthNonsizeIncreasing2002}. Like our stack-based tape, the array-based tape requires diamonds to interface with it, but it also carries around an index representing what element of the array is currently beneath the head of the tape. This index requires additional diamonds to store, leading to additional complexities in the implementation to properly share the total $n$ diamonds between the index and the tape operations.

As promised earlier, we can now combine all our results to obtain the following more general completeness theorem:
\begin{theorem}[Polynomial-Time Completeness]
\label{thm:completeness-general}
Let $A$ be a type such that $\dfree{A}$ and let $P : \mathbb N \to \mathbb N$ be a polynomial with degree $k$. For every $P$-time computable function $f : \tpsem{A}^* \to \tpsem{\tpsum{\tpunit{}}{A}}^*$, there exists a correct $k$-stack implementation with element type $\tpsum{\tpunit{}}{A}$, say with underlying type $S$ and items functions $I_n : \tpsem{S} \to \tpsem{\tpsum{\tpunit{}}{A}}^*$, and a closed term $M : \tparr{\tplist{A}}{S}$ such that for every $x \in \tpsem{A}^*$:
$$f(x) = I_{|x|}\left(\tmsem{M}{\envnil}(x)\right)$$
In other words, $M$ outputs a stack with items equal to $f(x)$.
\end{theorem}
\begin{proof}
To define $M$, we start with a function abstraction taking in the input $x : \tplist{A}$. Using the tape data structure discussed above, and the $n$ diamonds from $x$, we can write the values of $x$ onto the tape. Then, using the tools from \pref{sec:completeness-poly-iter}, we take an \lfpl implementation of the Turing machine's transition function and iterate it on this initial tape (as well as an \lfpl encoding of the initial machine state $q_0 \in \tpsem{Q}$) for a total of $P(n)$ steps. Finally, we return the stack representing the right half (or left; this choice just depends the irrelevant particulars of how we define a Turing machine's output) of the tape data structure.
\end{proof}

By allowing $M$ to output a stack rather than insisting it be a list, we can implement \textit{any} polynomial-time computation, even the size-increasing ones. Since the stack operations from \pref{sec:completeness-bounded-stack-definition} and the polynomial iterator from \pref{lem:polynomial-iterator} do not permanently consume any diamonds, we still have $n$ diamonds left over after the procedure in the above proof. We can just use those to remove the first $n$ values of the stack and store them in a list, thereby proving \pref{thm:completeness-main} as a corollary of \pref{thm:completeness-general}.