\section{Mechanization}
\label{sec:mechanization}

We have mechanized all results stated in Sections \ref{sec:lfpl}, \ref{sec:completeness}, and \ref{sec:soundness} in the proof assistant Istari \cite{istariProofAssistant}. To our knowledge, our mechanization serves as one of two published case studies in this relatively new proof assistant~\cite{Li2026}. The mechanization is roughly 7,000 lines of meaningful Istari code. Around 3,000 of these are definitions and \lfpl code, around 3,500 are proof scripts, and around 800 are lemma and theorem statements~\cite{mechanization}.

\subsection{Defining LFPL}

We begin by defining the types of \lfpl as an inductive datatype \lstinline{tp} within Istari. \lfpl does not have type variables or similar complexities. We also define the judgement $\dfree{A}$ within Istari. Next, we represent contexts $\ctx$ as an Istari list of \lfpl types. We do not need variable names associated with the types because we use De Bruijn indices in the term structure. Importantly, we define a judgement $\code{split} \; \ctx \; \ctx_1 \; \ctx_2$ to indicate when $\ctx$ can be  bipartitioned into $\ctxappend{\ctx_1}{\ctx_2}$, used extensively in the typing rules. Essentially, viewing the contexts as type lists as we do in our mechanization, our definition of $\code{split} \; \ctx \; \ctx_1 \; \ctx_2$ ensures $\code{append}(\ctx_1)(\ctx_2)$ is equivalent to some permutation of $\ctx$.

Instead of presenting an untyped syntax with an extrinsic typing judgement, as we do in \pref{sec:lfpl-definition}, our mechanization defines \lfpl's term syntax and typing rules at once, so that all terms are intrinsically well-typed under a known context. These intrinsically typed terms are represented by an inductive datatype \lstinline{term : ctx -> tp -> type}, where \lstinline{type} refers to Istari types, so that a value of type \lstinline{term G A} corresponds to a term $M$ such that $\types{G}{M}{A}$. For example, the rule for the term $\tmpair{M_1}{M_2}$ is:
\begin{lstlisting}
| pair : forall (G GA GB : ctx) (A B : tp) .
    split G GA GB -> term GA A -> term GB B -> term G (tensor A B)
\end{lstlisting}
It says we must bipartition the context $G$ into $\ctxappend{G_A}{G_B}$, and give a well-typed term under each of those contexts to form a pair $\tmpair{M_1}{M_2}$ which has type $\tptens{A}{B}$ under context $G$.



\subsection{Denotational and Operational Semantics}

Instead of the set-theoretic denotational semantics we give in \pref{sec:denotational-semantics}, our mechanization gives a denotational semantics from intrinsically typed \lfpl terms into Istari terms of the appropriate type. This is similar to the set-theoretic formulation, since all operations we use on sets (products, coproducts, etc.) are also present in the Istari type system. One benefit is that we can rely on Istari's reduction engine to simplify Istari terms resulting from the semantics of a complicated \lfpl term. These denotations have the following signatures.
Here, a value of type \lstinline{env G} represents an environment $\env$ such that $\env : G$. 
\begin{lstlisting}
tp_sem : tp -> type
term_sem : forall (G : ctx) (A : tp) . term G A -> env G -> tp_sem A
\end{lstlisting}
Similarly, for the operational semantics judgement
$\evals{\env}{M}{c}{v}$ from \pref{sec:operational-semantics}, we
define an inductive datatype encoding its rules with the following
type:
\begin{lstlisting}
evals : forall (G : ctx) (A : tp) . term G A -> env G -> value A -> nat -> type
\end{lstlisting}
As briefly mentioned at the end of \pref{sec:operational-semantics}, we give a denotational semantics for values and prove a coherence result (\pref{thm:coherence}) in our mechanization, which is a good example of what many of our Istari theorems look like.
\begin{lstlisting}
value_sem : forall (A : tp) . value A -> tp_sem A
value_sem = ...
lemma "operational_equiv_denotational"
  forall (G : ctx) (A : tp) (M : term G A) (V : env G) (v : value A) (c : nat) .
  evals M V v c -> value_sem v = term_sem M (env_sem V)
\end{lstlisting}




\subsection{Takeaways from Istari}

When beginning this project, we were uncertain which proof assistant would be most suitable for the mechanization, so we began by developing two mechanizations in parallel, one in Agda and one in Istari. During preliminary work for the completeness theorem, we ran into two significant difficulties with Agda.

The first issue was a lack of tactics. Agda expressions are often cleaner and more readable than the proof scripts of tactic-based proof assistants like Rocq and Istari. However, when one constantly needs to reason about equality, say by substituting $e'$ for $e$ in a term $f(e)$ where $f$ is a complex expression, it becomes helpful to have a substitution tactic. In Agda, we needed to manually specify $f$, which is infeasible when $f$ is large.

Secondly, Agda requires that all expressions are well-typed, even in the statement of a theorem yet to be proven. This presents some difficulties. For example, consider the \lfpl type $(\tplist{\tpunit})^k$ used in \pref{sec:completeness-bounded-stack-definition}. We know that $\tpsem{(\tplist{\tpunit})^k} = \tpsem{\tplist{\tpunit}}^k$, so we implicitly switch between these forms as needed, but in the mechanization this must be explicit. Some parts of our proof are only well-typed if they are given the form on the left-hand side of the equality, and others need the right-hand form. In Agda, we had to state a theorem where some parts of the theorem needed the left form, and others needed the right, so there was no way to formulate this such that it was well-typed without using transport functions, which are known to cause other major difficulties~\cite{tabareau2018}. Istari does not require the well-typedness of a theorem before it is proven. It must be given a type by the end of the proof, but this is not difficult since we have a lemma stating $\tpsem{(\tplist{\tpunit})^k} = \tpsem{\tplist{\tpunit}}^k$. The difference is that Istari lets us show well-typedness during the proof, whereas Agda demands it before the proof.

Such issues led us to finish the completeness and soundness proofs only in Istari. Retrospectively, we can say that it was a more than adequate tool for this task. In fact, we never encountered any Istari-related roadblocks. This is not to say there weren't annoyances; as a relatively new proof assistant, Istari sometimes struggles to infer implicit arguments and automatically eliminate vacuous proof cases, requiring manual intervention in situations where Agda would succeed.

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "../paper"
%%% End:
