\section{Polynomial-Time Soundness}
\label{sec:soundness}

This section shows that \lfpl programs can be evaluated in polynomial
time.
%
To strengthen the soundness result and illustrate how to add features to \lfpl while maintaining polynomial-time soundness, we extend the language and refer to the result as \lfplplus. The features are the standard product type, which manifests as the lazy product in an affine setting, a type of unbounded stacks which are similar to lists but do not support recursion, and the inductive type of binary trees. The syntax and typing rules of these language constructs are given in \pref{fig:ext-syntax-tp-tm} and \pref{fig:ext-jmt-ty} respectively.

\input{figures/ext-syntax-tp-tm}
\input{figures/ext-jmt-ty}

The typing rules for stacks of type $\tpstack{A}$ correspond to those for lists in a standard non-affine functional programming language, where the elimination form is case analysis rather than structural recursion and the introduction form does not require diamonds.
%
These stacks do not violate \lfpl's core principle of controlling recursion by diamonds because they do not support recursion.

The typing rules for the tree type $\tptree{A}$, an inductive type which does support recursion, are similar to those of lists $\tplist{A}$. The primary difference is that, in \pref{rule:ty-TreeE}, the base case $N_1$ is typed under an empty context. Like the list recursor, the tree recursor has two cases: the inductive and base case. The difference is that for the list recursor, the base case $N_1$ is only run once, so it may have access to part of the context (though this is not necessary for completeness). In the tree recursor, $N_1$ may be run many times, as a tree can have many leaves. So, to respect linearity, it cannot use anything from the context.

The denotational semantics for these new constructs are relatively straightforward. We set $\tpsem{\tpprod{A}{B}} = \tpsem{A} \times \tpsem{B}$ and $\tpsem{\tpstack{A}} = \tpsem{A}^*$. Giving a semantics for $\tptree{A}$ and its related terms boils down to giving a mathematical account of binary trees (in the same manner we have used the Kleene star to give a simple mathematical account of lists).
%
The semantics for terms involving these types can be found in the mechanization (see \pref{sec:mechanization}).

The stack type $\tpstack{A}$ is a particularly powerful addition to \lfpl. The vast majority of the difficulty in proving polynomial-time completeness of \lfpl, and by far the most complex aspect of our entire mechanization, was the definition and implementation of the bounded stack data structure discussed in \pref{sec:completeness-bounded-stack-definition} and \pref{sec:completeness-bounded-stack-implementation}. So, the addition of unbounded, $\tpdiam$-free stacks to \lfpl means \lfplplus admits a much simpler completeness proof. In essence, \lfplplus internalizes the bounded stack data structure from \pref{sec:completeness-bounded-stack-definition} and removes the boundedness restriction. No diamonds are required to interface with the stacks of \lfplplus, and there is no upper bound on the number of stack elements. The addition of a feature like this to \lfpl would eliminate the need for the hardest part of our completeness proof. \lfplplus also contains some other generally useful features to showcase our method of proving soundness, such as lazy products and binary trees.



\subsection{Big-Step Operational Cost Semantics}
\label{sec:operational-semantics}

To reason about the cost of evaluating \lfplplus terms, we need to reason about evaluating functions which may take lists as input. For example, we have seen a closed \lfpl term $\code{reverse} : \tparr{\tplist{A}}{\tplist{A}}$, but we can never actually ``run'' $\code{reverse}$ on a nonempty input list, since that would require a closed term of type $\tpdiam$ and the whole point of \lfpl's type system is to make that impossible.

To solve this problem, we introduce a separate language for \lfplplus values (see \pref{fig:syntax-val}). Note the definitions of value and environment are mutually recursive. This means we usually have to define functions on values via mutual recursion with a corresponding function on environments. Also, note that there is a value $\vdiam$ of type $\tpdiam$. Like in the denotational semantics, diamonds in lists and trees are not explicit.
%
By separating values from terms, we ensure \lfplplus terms remain polynomial-time sound while obtaining a means of providing concrete inputs to observe their evaluation. We also define a typing judgement $v : A$ for values, so that $\vdiam : \tpdiam$, $\vcons{v_h}{v_t} : \tplist{A}$ whenever $v_h : A$ and $v_t : \tplist{A}$, and so on.

\input{figures/syntax-val}

To account for the cost of evaluation, we give a cost-annotated big-step operational semantics, represented by the judgement $\evals{\env}{M}{c}{v}$, read ``term $M$ evaluates under the environment $\env$ to a value $v$ with cost $c$''. A representative set of rules for this judgement is given in \pref{fig:jmt-eval}, and a complete set of rules can be found in{
  \ifdefined\FULLVERSION
    \pref{app:evaluation}
  \else
    Appendix A of this paper's full version~\cite{fullversion}
  \fi
}. Note the presence of constants $\Cvar, \Cnull, \ldots, \Ctrec \in \mathbb N$ (one for each syntactic construct) in the conclusion of each rule. These are left arbitrary so that our cost model is generic with respect to the particular costs of each primitive operation of the language.

This semantics enjoys several standard properties such as determinism and preservation of typing, which we state here and prove in the mechanization.

\begin{theorem}[Determinism]
If $\evals{\env}{M}{c}{v}$ and $\evals{\env}{M}{c'}{v'}$ then $c = c'$ and $v = v'$.
\end{theorem}

\begin{theorem}[Preservation]
If $\types{\Gamma}{M}{A}$, $\envtypes{\env}{\Gamma}$, and $\evals{\env}{M}{c}{v}$, then $v : A$.
\end{theorem}
We can also give a denotational semantics for values and environments, and prove that they cohere with the denotational semantics for terms. In particular, given $v : A$ and $\envtypes{\env}{\Gamma}$, we define $\tpsem{v} \in \tpsem{A}$ and $\tpsem{\env} \in \tpsem{\Gamma}$ by mutual induction on the structures of $v$ and $\env$. Most cases are straightforward; for example, $\tpsem{\vpair{v_1}{v_2}} = (\tpsem{v_1}, \tpsem{v_2})$. For closures such as function values, we appeal to the term denotational semantics: $\tpsem{\vlam{\env}{x}{M}} = \lambda v . \tmsem{M}{\envcons{\tpsem{\env}}{x}{v}}$. We prove the following in the mechanization:

\begin{theorem}[Coherence]
If $\types{\Gamma}{M}{A}$, $\envtypes{\env}{\Gamma}$, and $\evals{\env}{M}{c}{v}$, then $\tpsem{v} = \tmsem{M}{\tpsem{\env}}$.
\end{theorem}

Polynomial-time soundness is meant to be a statement about the computational complexity (of functions encoded by \lfpl) with respect to some machine model like a Turing machine. The big-step cost semantics is somewhat removed from a Turing machine's cost model.
%
However, it is well understood how to justify a cost semantics for a functional language like \lfplplus in terms of a low-level model such as a Turing machine \cite{BlellG96,Accattoli2019}.
%
The only non-standard features of \lfpl are the affine type system and the type $\tpdiam$.
%
However, both of these can be forgotten in a simple cost-preserving translation.
%
Thus, when proving true polynomial-time soundness of \lfpl, the only part which relies on \lfpl-specific reasoning is the very first step of giving a polynomial bound on the cost in \lfpl's big-step semantics; after that, these standard methods will carry the polynomial bound all the way down to the machine cost model. As the focus of this paper is \lfpl, we only mechanize the construction and verification of the polynomial cost bound.

\input{figures/jmt-eval}



\subsection{The Non-Size-Increasing Property}
\label{sec:soundness-size}

We now make precise the non-size-increasing property referred to in previous sections. Our goal is to assign to each value $v$ a number $\valsize{v} \in \mathbb N$ that, roughly speaking, counts the number of diamond values $\vdiam$ contained within $v$.
%
We then show that the size of the value resulting from the evaluation of an \lfplplus term is bounded by the size of the values in the environment. We define size by mutual induction on values and environments in \pref{fig:size-val-env}.

\input{figures/size-val-env}

% Our intuitive notion of the non-size-increasing property is that a function's output should have size no greater than its input.
\begin{theorem}[The Non-Size-Increasing Property]
\label{thm:non-size-increasing}
Suppose that $\types{\ctx}{M}{A}$. Then, for all environments $\env$, values $v$, and costs $c$ such that $\evals{\env}{M}{c}{v}$, we have $\valsize{v} \le \envsize{\env}$.
\end{theorem}
\begin{proof}
By rule induction on $\evals{\env}{M}{c}{v}$.
\end{proof}

The relevance of this result to soundness might already be clear. An immediate corollary of the theorem is that there are no closed terms of type $\tpdiam$, one of the main properties \lfpl's type system was designed to ensure. The reason is that $\vdiam$ has size $1$ but the empty environment has size $0$.



\subsection{Polynomial Cost Bound}

Our goal for the rest of this section is to prove the following:

\begin{theorem}[Concrete Polynomial-Time Soundness of \lfplplus]
\label{thm:soundness-concrete}
Suppose $\types{\ctxcons{\ctxnil}{\ell}{\tplist{A}}}{M}{B}$, where $\dfree{A}$. Then, there exists a polynomial $P : \mathbb N \to \mathbb N$ such that, for any value $v_\ell : \tplist{A}$ representing a list of length $n$, there exists a value $v : B$ and a cost $c \in \mathbb N$ such that $\evals{\envcons{\envnil}{\ell}{v_\ell}}{M}{c}{v}$ and $c \le P(n)$.
\end{theorem}

Note it is important that the polynomial only depends on $M$, i.e. it is quantified before the input $v_\ell$. Taking $B = \tplist{\tpunit}$ and $A = \tpunit$, we recover the basic notion of soundness for \lfplplus functions on natural numbers. We present this more general form to motivate our transition to the even more general form we need to successfully prove it. As usual, we must reason about arbitrary terms under arbitrary contexts. It is tempting to say the following:

\begin{center}
\parbox{.9\linewidth}{
  \it
Suppose $\types{\ctx}{M}{A}$. There exists a polynomial $P : \mathbb N \to \mathbb N$ such that, if $\env : \Gamma$, there exists a value $v : A$ and a cost $c \in \mathbb N$ such that $\evals{\env}{M}{c}{v}$ and $c \le P(\envsize{\env})$.
}
\end{center}

However, this statement is false. The problem is that the environment may contain values which incur cost when used, which is dynamic information and $P$ only has access to static information. Consider $\types{f : \tparr{\tpunit}{\tpunit}}{\tmapp{f}{\tmnull}}{\tpunit}$. If for simplicity $\Capp = \Cvar = \Cnull = 0$, we cannot statically observe how this program could incur any cost. After all, the only statically observable syntactic constructs are application, variable, and null. On the other hand, dynamically, $\env$ could map $f$ to a function which incurs cost.

To resolve this issue, we must consider the potential cost of running functions (or lazy pairs) stored in the environment $\env$ and result value $v$. Suppose for a moment we have defined three families of polynomials: $P_M$ on all terms $M$, $P_v$ on all values $v$, and $P_\env$ on all environments $\env$. Our new goal is to prove the following theorem.

\begin{theorem}[General Soundness]
\label{thm:soundness-general}
Suppose $\types{\ctx}{M}{A}$. Then, for any $\env : \Gamma$, there exists a value $v : A$ and a cost $c \in \mathbb N$ such that $\evals{\env}{M}{c}{v}$ and:
$$c + P_v(\envsize{\env}) \le P_M(\envsize{\env}) + P_{\env}(\envsize{\env})$$
\end{theorem}

This theorem is true (for some definitions of $P_v$, $P_M$, and $P_\env$), though as stated it is not strong enough to be proven by rule induction on $\types{\ctx}{M}{A}$. Before discussing this further, we define the polynomials used in this theorem. By $\max(P,Q) : \mathbb N \to \mathbb N$ we mean the polynomial with coefficients equal to the pairwise maximums of the coefficients of polynomials $P,Q : \mathbb N \to \mathbb N$. This way, $\max(P(n), Q(n)) \le \max(P,Q)(n)$.

\begin{definition}[Term Polynomial]
Let $M$ be a \lfplplus term. We inductively define the polynomial $P_M : \mathbb N \to \mathbb N$ as follows.
\upshape{
\begin{align*}
P_{x}(n) &= \Cvar \\
P_{\tmrecord{M_1}{M_2}}(n) &= \Crecord + \max(P_{M_1}, P_{M_2})(n) \\
P_{\tmproj{i}{M}}(n) &= \Cproj{i} + P_M(n) \\
P_{\tmlam{x}{M}}(n) &= \Clam + P_M(n) \\
P_{\tmapp{M}{N}}(n) &= \Capp + P_M(n) + P_N(n) \\
P_{\tmcase{M}{x_1}{N_1}{x_2}{N_2}}(n) &= \Ccase + P_M(n) + \max(P_{N_1}, P_{N_2})(n) \\
P_{\tmrec{M}{N_1}{x_d}{x_h}{x_t}{N_2}}(n) &= P_M(n) + (\Crec + P_{N_1}(n)) \\
&\quad\quad + n (\Cvar + \Crec + P_{N_2}(n)) \\
P_{\tmtrec{M}{N_1}{x}{y}{z}{w}{N_2}}(n) &= P_M(n) + (n + 1) (\Ctrec + P_{N_1}(n)) \\
&\quad\quad + n (2\Cvar + \Ctrec + P_{N_2}(n))
\end{align*}
}
The remaining cases are similar and provided in the mechanization.
\end{definition}

The aim of $P_M$ is to bound the cost of evaluating $M$, and the future cost of using any closures it evaluates to, by $P_M(n)$ in an environment containing $n$ diamonds. The intuition behind $P_{\tmrec{M}{N_1}{x_d}{x_h}{x_t}{N_2}}(n)$ is that $M$ and $N_1$ are evaluated once, and $N_2$ could be evaluated up to $n$ times, where $n = \envsize{\env}$ is the number of diamonds available the environment. Similarly, for $P_{\tmtrec{M}{N_1}{x}{y}{z}{w}{N_2}}$, the base case $N_1$ could be evaluated $n + 1$ times (a tree with $n$ internal nodes has $n + 1$ leaf nodes) and $N_2$ could be evaluated $n$ times.

\begin{definition}[Value and Environment Polynomials]
  We define polynomials $P_\env : \mathbb N \to \mathbb N$ and
  $P_v : \mathbb N \to \mathbb N$ via mutual induction as follows
\begin{align*}
P_{\vdiam} &= P_{\vnull} = P_{\vempty} = P_{\vnil} = P_{\vleaf} = 0 \\
P_{\vrecord{\env}{M_1}{M_2}} &= P_{\env} + \max(P_{M_1}, P_{M_2}) \\
P_{\vlam{\env}{x}{M}} &= P_{\env} + P_M \\
P_{\vpair{v_1}{v_2}} &= P_{v_1} + P_{v_2} \\
P_{\envnil} &= 0 \quad\text{and}\quad P_{\envcons{\env}{x}{v}} = P_\env + P_v
\end{align*}
The definition of $P_v$ for other values $v$ is given in the mechanization.
\end{definition}
The idea of  $P_v$ and $P_\env$ is to bound the future cost of evaluating all of the closures (functions and lazy pairs) stored inside $v$ and $\env$ respectively. With these choices of polynomials, \pref{thm:soundness-general} implies \pref{thm:soundness-concrete}, because when $v : \tplist{A}$ and $\dfree{A}$, it is clear that $P_v = 0$ and $\valsize{v} = n$, where $n$ is the length of the list represented by $v$.



\subsection{Proving the Soundness Theorem}
\label{sec:proving-general-soundness}

To prove \pref{thm:soundness-general}, we take a standard logical relations approach. We define a unary logical relation $\mathcal R$ for terms, values, and environments with the goal that the fundamental theorem for $\mathcal R$ should imply \pref{thm:soundness-general}.

\begin{definition}
We first define the shorthand $\Rterm{A}{\env}{M}$ to mean:
\begin{align*}
&\text{For all $n \ge \envsize{\env}$, there exists $v : A$ and a cost $c \in \mathbb N$ such that $\evals{\env}{M}{c}{v}$,} \\
&\text{$\Rval{A}{v}$ holds, and $c + P_v(n) \le P_M(n) + P_\env(n)$}
\end{align*}
\end{definition}
By working with a general $n \ge \envsize{\env}$ rather than just $\envsize{\env}$ as the input to the polynomials, we can deal with the case where $\env$ arises as part of a larger environment during the proof. This is reminiscent of ``future worlds'' from Kripke-style logical relations. % TODO: Citation?

\begin{definition}
We define $\Rval{A}{v}$ by induction on the type $A$:
\begin{align*}
\Rval{\tpunit}{v} &\quad \text{iff} \quad v = \vnull \\
\Rval{\tptens{A_1}{A_2}}{v} &\quad \text{iff} \quad v = \vpair{v_1}{v_2}, \; \Rval{A_2}{v_1}, \text{ and } \Rval{A_2}{v_2} \\
\Rval{\tptens{A_1}{A_2}}{v} &\quad \text{iff} \quad v = \vpair{v_1}{v_2}, \; \Rval{A_2}{v_1}, \text{ and } \Rval{A_2}{v_2} \\
\Rval{\tparr{A}{B}}{v} &\quad \text{iff} \quad v = \vlam{\env}{x}{M}, \text{ and } \Rval{A}{v} \text{ implies } \Rterm{B}{\envcons{\env}{x}{v}}{M} \\
\Rval{\tplist{A}}{v} &\quad \text{iff} \quad \text{$v$ is a list $\ell$ such that } \Rval{A}{-} \text{ holds for each element of $\ell$ }
\end{align*}
The remaining cases are provided in the mechanization.
\end{definition}

\begin{definition}
We define $\Renv{\Gamma}{\env}$ by induction on $\Gamma$. $\Renv{\ctxnil}{\envnil}$ always holds. $\Renv{\ctxcons{\Gamma}{x}{A}}{\envcons{\env}{x}{v}}$ holds whenever $\Rval{A}{v}$ and $\Renv{\Gamma}{\env}$ hold.
\end{definition}

Now, \pref{thm:soundness-general} is clearly implied by the fundamental theorem for our relation $\mathcal R$:

\begin{theorem}[Fundamental Theorem]
Suppose $\types{\Gamma}{M}{A}$ and $\envtypes{\env}{\Gamma}$. Then, if $\Renv{\Gamma}{\env}$ holds, so does $\Rterm{A}{\env}{M}$.
\end{theorem}
\begin{proof}
We go by rule induction on $\types{\ctx}{M}{A}$. \pref{thm:non-size-increasing} comes up in nearly every case, since we need to satisfy the premise $n \ge \envsize{\env}$ of $\Rterm{A}{\env}{M}$ to use any inductive hypotheses. Importantly, we actually make use of this premise in the list and tree recursor cases, where we can say for example that the length of any list value $v$ is at most $\envsize{\env}$ and therefore at most $n$. Most cases are verbose but straightforward, and a couple of the less straightforward cases can be found in{
  \ifdefined\FULLVERSION
    \pref{app:soundness}
  \else
    Appendix B of the full version~\cite{fullversion}
  \fi
}.
\end{proof}